---
title: "Example workflow for ceramic analysis using potteR"
author: "Nick Gestrich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ceramic_analysis_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(potteR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Example Data

This example runs on data from a 2023 excavation in the Republic of Guinea. These are recorded using a modified version of the system proposed by Jesse & Nowotnick (2022). It is included in the package.

```{r}
rims <- guinea[[1]]
body <- guinea[[2]]

#some of the observations on rims are not made on body sherds. We will treat these observations separately
all <- guinea[[3]]

r <- names(rims)[! names(rims) %in% names(body)]

rims_only <- rims |> 
  select(all_of(r), site, sondage, context, unit_id, sond_con)
```

## 1. Univariate summary statistics

We begin our explorative data analysis with a report of summary statistics and counts of each variable. This means counts of unique values in the categorical variables, and summaries of the numeric variables. PotteR provides these as an R object in a list, and as a pdf report for perusal outside of R. The pdf version also includes a basic visualisation for each variable.

```{r}
# make a list of summaries for every variable
rim_summary <- get_summary(rims_only)

# to generate a pdf report, uncomment and run the code below
print_summary(rims_only, file = here::here("summary.pdf"))
```

It will often not be necessary to run these functions on the entire dataset. In that case, a selection of variables can be made first.

```{r}
#| output: false
rims_only |> 
  select(parts:rim_type) |> 
  get_summary()
```

## 2. Fixing errors in the data

The univariate reports frequently show up errors in the data. These should be fixed and the reports rerun before moving on to further analyses.

### Replace individual values

The package provides a small wrapper function to replace individual cases.

```{r}
replace_values <- function(data, col, old, new){
  data |> 
    dplyr::mutate({{col}} := replace({{col}}, {{col}} == old, new))
}

rims_only <- rims_only |> replace_values(rim_angle, 43, 4)
```

Note that this method does not apply to NA values.

### Setting and replacing NA values

To replace all NAs in a variable with a different value (e.g. 0), use the replace_na() function from the tidyr package.

```{r}
rims_only$rim_angle |> replace_na("missing")
```

To replace a value with NA, use the na_if() function from the dplyr package. Also see the documentation [here](https://dplyr.tidyverse.org/reference/na_if.html).

```{r}
rims_only |> mutate(diam = na_if(diam, 0))

# do this for several columns using across()
rims_only |> 
  mutate(across(where(is.numeric), ~na_if(., 0)))
```

### Making several replacements

More numerous replacements are best done using dplyr's case_match() function in the following manner:

```{r}
rims_only <- rims_only |> 
  dplyr::mutate(rim_type = case_match(rim_type, 
                                      c("SI", "Si") ~ "S1",
                                      c("T1(?)", "T1...") ~ "T1",
                                      .default = rim_type
                                      ))
```

## 3. Bivariate statistics

The next step in the exploratory phase of the analysis is to survey the variables for possible correlation. It is expedient to do this visually at first. The package provides a function which generates a pdf of bivariate combinations. Look at these and decide which relationships to follow up on.


```{r}
# Function that creates one of three plots, according to variable type:
make_quick_plot <- function(data, x_var, y_var, ...){
  
  plot_box <- function(df, catvar, numvar, ...){
  df |> 
    select({{catvar}}, {{numvar}}) |> 
    tidyr::drop_na() |> 
    ggplot(aes(x = reorder({{catvar}}, {{numvar}}), y = {{numvar}})) + 
    geom_boxplot()+
    theme_bw()+
    labs(x = rlang::englue("{{catvar}}"),
         y = rlang::englue("{{numvar}}"))
  }
  
  plot_bar_group <- function(df, x, y, ...){
  df |>
    select({{x}}, {{y}}) |> 
    tidyr::drop_na() |> 
    ggplot(aes(x = {{ x }}, fill  = {{ y }}))+
    geom_bar(position = "dodge")+
    theme_bw()
  }
  
  
  if(is.numeric(eval(substitute(x_var), data)) && is.numeric(eval(substitute(y_var), data))){
  
    data |> 
    ggplot(aes(x = {{x_var}}, y = {{y_var}})) + 
    geom_point()+
    geom_smooth(method = "lm")+  
    theme_bw()

  }
  
  else if(is.character(eval(substitute(x_var), data)) && is.numeric(eval(substitute(y_var), data))){

    
    plot_box(data, catvar = {{x_var}}, numvar = {{y_var}})
    
  }
  
  else if(is.character(eval(substitute(y_var), data)) && is.numeric(eval(substitute(x_var), data))){

    
    plot_box(data, catvar = {{y_var}}, numvar = {{x_var}})
    
  }
    
  else if(is.character(eval(substitute(x_var), data)) && is.character(eval(substitute(y_var), data))){
    if(n_distinct(eval(substitute(x_var), data)) > n_distinct(eval(substitute(y_var), data))){
      plot_bar_group(data, {{x_var}}, {{y_var}})
    }
    else{
      plot_bar_group(data, {{y_var}}, {{x_var}})
    }
  }
}


rims_only |> make_quick_plot(mx_th, rim_angle)



# it might be a good idea to center and standardise numerical values before analysing their distribution. To discuss with Cezary.


center_and_standardise <- function(x) (x - mean(x, na.rm = TRUE) / sd(x, na.rm = TRUE))


library(ggplot2)

l <- combn(names(rims_only), 2, simplify = FALSE)

l <- purrr::map(l, append, 'rims_only', after = 0)



p <- do.call(make_quick_plot, as.list(l[[1]]))

purrr::map(l, ~do.call(make_quick_plot, as.list(.x)))
ldf <- as.data.frame(do.call(rbind, l))
make_quick_plot(data = eval(parse(text = ldf$V1)), ldf$V2, ldf$V3)

lapply(l, function(x) do.call(make_quick_plot, as.list(x)))

x <- as.list(ldf$V2)
y <- as.list(ldf$V3)

purrr::map2(x, y, make_quick_plot, data = rims_only)

out <- vector("list", length = length(l))
for(i in seq_along(l)){
  
 out[[i]] <-  rims_only |> 
   ggplot(aes_string(x = l[[i]][1], y = l[[i]][2])) + 
  geom_point()+
  geom_smooth(method = "lm")+  
  theme_bw()
}
out[[20]]


plot_scatter <- function(data, x_var, y_var){
  data |> 
  ggplot(aes(x = .data[[x_var]], y = .data[[y_var]])) + 
  geom_point()+
  geom_smooth(method = "lm")+  
  theme_bw()
}



p <- names(rims_only)[-1] |> 
  purrr::map( ~ make_quick_plot(rims_only, names(rims_only)[1], .x))

p <- l |> 
  purrr::map( function(x) make_quick_plot(rims_only, l[[x[1]]], l[[x[2]]]))

p[[20]]
print_2var <- function(data){}

print(p)
```


Do correlation tests, then order the plots according to correlation 


## Plots

We provide a ggplot theme for fast and unified styling of graphics.


```{r}
ggplot(rims) +
  geom_point(aes(x = diam, y = mx_th, colour = sondage)) +
  theme_potteR()

```



```{r}
ggplot(rims) +
  geom_boxplot(aes(x = sondage, y = mx_th, fill = sondage)) +
  params_potteR

params_potteR <- list(ggplot2::theme_bw(), 
                      khroma::scale_colour_bright())
theme_potteR <- function() {
  ggplot2::theme_bw()+
  khroma::scale_colour_bright()
}
  
```




## Maps

## Seriation

An important aspect of ceramic analysis is to find attributes or types that seriate. This has two main interpretive ends in archaeology. The first is to create orders in previously unordered data, for instance in a series of samples from survey or shovel testing. The second is to identify developments in series that have pre-defined orders, such as contexts from stratigraphic excavations.

### Using seriation to order assemblages

### Identifying developments in ordered assemblages

We will here aim our seriation plot at decor types and rim types. We use the `plot_ford()` function from the [`tabula`](https://www.tesselle.org/packages.html) package to make these plots. The function given here adds ease in data preparation.

A predefined order has to be given for the contexts or assemblages. This is then used as the `context_order` argument in the function.

```{r}
# Stratigraphic order of contexts for Unit B
b_order <- c("1", "2", "3", "4", "5",  "6", "7",  "8",  "9",  "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23",  "24", "25", "26")

rims_only |> 
  filter(sondage == "B") |> # select only Unit B
  plot_series(context, rim_type, b_order)
```

Since the returned plot is a ggplot object, it can be modified using ggplot syntax to add titles etc., e.g.:

```{r}
rims_only |> 
  filter(sondage == "B") |> # select only Unit B
  plot_series(context, rim_type, b_order) +
  labs(title = "Rim Types from Unit B",
       subtitle = "NDK1")
```

## Finding Groups (Juan-Marco)

## Analysing decors and motifs
networks (J-M)



## Significance and Standardisation


