---
title: "Example workflow for ceramic analysis using potteR"
author: "Nick Gestrich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ceramic_analysis_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(potteR)
library(dplyr)
```

## Example Data

This example runs on data from a 2023 excavation in the Republic of Guinea. These are recorded using a modified version of the system proposed by Jesse & Nowotnick (2022). It is included in the package.

```{r}
rims <- guinea[[1]]
body <- guinea[[2]]

#some of the observations on rims are not made on body sherds. We will treat these observations separately
all <- guinea[[3]]

r <- names(rims)[! names(rims) %in% names(body)]

rims_only <- rims |> 
  select(all_of(r), site, sondage, context, unit_id, sond_con)
```

## Summary statistics

We would like a report table of key statistics and counts. This means counts of unique values in the categorical variables, and summaries of the numeric variables

```{r}
xx <- get_summary(rims_only)

get_summary <- function(data) {
  x <- data |> sum_categoricals()
  y <- data |> sum_numerics()
  x <- append(x, list(y), 0)
  names(x)[1] <- "numericals"
  return(x)
}




```
