---
title: "Example workflow for ceramic analysis using potR"
author: "Nick Gestrich"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ceramic_analysis_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
# delete before publication
library(devtools)
load_all()
```


```{r setup, warning=FALSE, message=FALSE}
library(potteR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Example Data

This example runs on data from a 2023 excavation in the Republic of Guinea. These are recorded using a modified version of the system proposed by Jesse & Nowotnick (2022). It is included in the package.

```{r}
rims <- guinea[[1]]
body <- guinea[[2]]

#some of the observations on rims are not made on body sherds. We will treat these observations separately
all <- guinea[[3]]

r <- names(rims)[! names(rims) %in% names(body)]

rims_only <- rims |> 
  select(all_of(r), site, sondage, context, unit_id, sond_con)
```

## 1. Univariate summary statistics

We begin our explorative data analysis with a report of summary statistics and counts of each variable. This means counts of unique values in the categorical variables, and summaries of the numeric variables. PotteR provides these as an R object in a list, and as a pdf report for perusal outside of R. The pdf version also includes a basic visualisation for each variable.

```{r}
# make a list of summaries for every variable
rim_summary <- get_summary(rims_only)

# to generate a pdf report, uncomment and run the code below
#print_summary(rims_only, file = here::here("summary.pdf"))
```

It will often not be necessary to run these functions on the entire dataset. In that case, a selection of variables can be made first.

```{r}
#| output: false
rims_only |> 
  select(parts:rim_type) |> 
  get_summary()
```

## 2. Fixing errors in the data

The univariate reports frequently show up errors in the data. These should be fixed and the reports rerun before moving on to further analyses.

### Replace individual values

The package provides a small wrapper function to replace individual cases.

```{r}
replace_values <- function(data, col, old, new){
  data |> 
    dplyr::mutate({{col}} := replace({{col}}, {{col}} == old, new))
}

rims_only <- rims_only |> replace_values(rim_angle, 43, 4)
```

Note that this method does not apply to NA values.

### Setting and replacing NA values

To replace all NAs in a variable with a different value (e.g. 0), use the replace_na() function from the tidyr package.

```{r}
rims_only$rim_angle |> replace_na("missing")
```

To replace a value with NA, use the na_if() function from the dplyr package. Also see the documentation [here](https://dplyr.tidyverse.org/reference/na_if.html).

```{r}
rims_only |> mutate(diam = na_if(diam, 0))

# do this for several columns using across()
rims_only |> 
  mutate(across(where(is.numeric), ~na_if(., 0)))
```

### Making several replacements

More numerous replacements are best done using dplyr's case_match() function in the following manner:

```{r}
rims_only <- rims_only |> 
  dplyr::mutate(rim_type = case_match(rim_type, 
                                      c("SI", "Si") ~ "S1",
                                      c("T1(?)", "T1...") ~ "T1",
                                      .default = rim_type
                                      ))
```

## 3. Split the data

## 4. Bivariates

## 5. Maps

## 6. Seriation

An important aspect of ceramic analysis is to find attributes or types that seriate. This has two main interpretive ends in archaeology. The first is to create orders in previously unordered data, for instance in a series of samples from survey or shovel testing. The second is to identify developments in series that have pre-defined orders, such as contexts from stratigraphic excavations.

### Using seriation to order assemblages

### Identifying developments in ordered assemblages

We will here aim our seriation plot at decor types and rim types. We use the `plot_ford()` function from the [`tabula`](https://www.tesselle.org/packages.html) package to make these plots. The function given here adds ease in data preparation.

A predefined order has to be given for the contexts or assemblages. This is then used as the `context_order` argument in the function.

```{r}
# Stratigraphic order of contexts for Unit B
b_order <- c("1", "2", "3", "4", "5",  "6", "7",  "8",  "9",  "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23",  "24", "25", "26")

rims_only |> 
  filter(sondage == "B") |> # select only Unit B
  plot_series(context, rim_type, b_order)
```

Since the returned plot is a ggplot object, it can be modified using ggplot syntax to add titles etc., e.g.:

```{r}
rims_only |> 
  filter(sondage == "B") |> # select only Unit B
  plot_series(context, rim_type, b_order) +
  labs(title = "Rim Types from Unit B",
       subtitle = "NDK1")
```


## Simplified rims (PS)
This function selects only the letters in front of the numbers in the rim_type. The aim is to only get a general type e.g. simple, so it is easier to categorise
This function should probably come up earlier in the script
```{r}
# load rim data
rims <- guinea[[1]]

# clean rim types
rims <- rims |> 
  dplyr::mutate(rim_type = dplyr::case_match(rim_type, 
                                      c("SI", "Si") ~ "S1",
                                      c("T1(?)", "T1...") ~ "T1",
                                      .default = rim_type
                                      ))

#add simple rimtypes
rims <- add_simple_rim(rims, column = "rim_type")

```


## Finding Groups based on FAMD (PS)
The function allows to search for groups of similar vessels based on selected variables. The hierarchical clustering (results saved at[[3]]) is based on the results of a Factor Analysis of Mixed Data (results saved at[[4]]). This clustering results in a column that is added to the data (to the whole dataset [[1]] or a dataset consisting only of the variables used in the FAMD [[2]]). If graph = TRUE the graphs produced with the FAMD aswell as a visualisation of the clusters in the 1/2-dimensions of the FAMD is provided.
In this example (which is also the default columns) diameter, rim_angle and simplified rimtypes are used to find groups of vessels based on their form.
```{r}


fg <- famd_groups(data = rims, column = c("diam", "rim_angle", "simple_rim_type"), graph = TRUE)

#taking a look at the result of this function
fg

#saving the modified data
rims_clust <- fg[[2]]

#inspecting the modified data
rims_clust |> 
  utils::head(n = 5)

```
```{r}
fg[[1]] <- fg[[1]] |> rename(clust_drs = `clust_ drs`)

fg[[1]] |> 
  as_tibble() |>
  filter(sondage == "B") |> # select only Unit B
  plot_series(context, clust_drs, b_order)
```



## Significance and Standardisation


