---
title: "Untitled"
author: "Nikolas Gestrich"
date: "2023-07-18"
output: 
  word_document:
  df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(devtools)
load_all()

library(potteR)
library(dplyr)
library(purrr)
library(knitr)
```

## Example Data

This example runs on data from a 2023 excavation in the Republic of Guinea. These are recorded using a modified version of the system proposed by Jesse & Nowotnick (2022). It is included in the package.

```{r}
rims <- guinea[[1]]
body <- guinea[[2]]

#some of the observations on rims are not made on body sherds. We will treat these observations separately
all <- guinea[[3]]

r <- names(rims)[! names(rims) %in% names(body)]

rims_only <- rims |> 
  select(all_of(r), site, sondage, context, unit_id, sond_con)
```

## Summary statistics

We would like a report of key statistics and counts. This means counts of unique values in the categorical variables, and summaries of the numeric variables
```{r}
rim_summary <- get_summary(rims_only)
```

```{r}
out <- purrr::map_chr(
    rim_summary,
    function(tt){
        knit_expand(text = "{{tt}}")
        }
    )
```

`r paste(knit(out), collapse = "\n")`
